---
title: "Headway Estimation"
author: "Tom Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidytransit-headways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidytransit)
```

# Introduction

This is a brief introduction to the functions in tidytransit that can be used to describe the frequency with which vehicles are scheduled to pass through routes and stops. 

# Getting headways using read_gtfs()

For convenience, when you pass a `frequency=TRUE` parameter to `read_gtfs()`, a `routes_frequency` dataframe is added to the list of calculated dataframes in the gtfs object as read by `read_gtfs`. 

## Key Assumptions
 
By default `read_gtfs` assumes:

- the user wants headways from 6 AM and 10 PM for all schedules that run on all weekdays.
- the user wants headways that correspond to the frequency with which vehicles pass through stops. 
- [that we can use a heuristic to guess which service_id is representative of a standard weekday](https://github.com/r-transit/tidytransit/blob/master/R/frequencies.R#L34-L59). 

See the reference for the `get_route_frequency` function for other options (e.g. weekends, other times of day). 

```{r}
local_gtfs_path <- system.file("extdata", 
                               "google_transit_nyc_subway.zip", 
                               package = "tidytransit")
nyc <- read_gtfs(local_gtfs_path, 
                 local=TRUE,
                 geometry=TRUE,
                 frequency=TRUE)
```

### Headways by Route

View the headways along routes as a dataframe. 

```{r}
head(nyc$.$routes_frequency)
```

### Headways by Stop

View the headways at stops. `stops_frequency` is added to the list of gtfs dataframes read in by `read_gtfs`. Again, by default, frequency is calculated for service that happens every weekday from 6 am to 10 pm. See the reference for the `get_stop_frequency` function for other options (e.g. weekends, other times of day). 

```{r}
head(nyc$.$stops_frequency)
```

### Plot Frequency Map

You can now map subway routes and color-code each route by how often trains come.  
```{r}
plot(nyc)
```

You can also use the `get_route_frequency` function directly, which takes arguments for other windows of time, and for other schedules. See the `get_route_frequency` documentation in the reference for more detail.

# Plotting Stop Headways

Before we plot headways at stops, we must join the frequency table to the geometries for the stops. 

```{r}
some_stops_freq_sf <- nyc$.$stops_sf %>%
  left_join(nyc$.$stops_frequency, by="stop_id") %>%
  select(headway)
```

Then we can plot them. 

```{r}
plot(some_stops_freq_sf)
```

We will see some outliers for headway calculations in this plot. 

In the NYC MTA schedule, for a few stops, a train will only show up a few times a day. Since we are calculating headways, by default, for a period from 6 am to 10 pm, the average headway for these stops will be as high as hundred of minutes. 

One quick solution to the outlier stops in above plot is to throw out stops with headways greater than an unreasonable amount of time. For example, we can filter out stops with headways above 60 minutes. 

```{r}
some_stops_freq_sf <- some_stops_freq_sf %>%
  filter(headway<60)
plot(some_stops_freq_sf)
```

If you're interested in how to work with schedules and outlier stops like this, the `timetable` vignette, included in this package, is a great introduction.  

### Route Frequencies

To calculate frequency at the route level, by default, tidytransit summary statistics of the calculated headways at stops along each route. So the median headways for stops along each route do some work to throw out outliers stop headways.  

One way to verify that the frequency and service calculations from GTFS are accurate is by checking against other sources, such as the train schedules themselves. 

```{r}
head(nyc$.$routes_frequency)
```

Above, we see that the median headway for the 1 train from 6 AM to 10 PM is 5 minutes according to our calculations. According to the [wikipedia entry](https://en.wikipedia.org/wiki/List_of_New_York_City_Subway_services#Train_intervals) for the NYC MTA this seems about right. 

#### Getting Frequencies for A Specific Day and Time

We might also want to check what rush hour headways are like on a specific day. The `set_hms_times` and `set_date_service_table` functions will alter the feed for us, allowing us to filter by date. 

```{r}
nyc <- nyc %>% 
  set_hms_times() %>% 
  set_date_service_table()
```

Below we pull a service ID for a specific weekday (2018-08-23). 

```{r}
nyc <- nyc %>% 
  set_hms_times() %>% 
  set_date_service_table()

services_on_180823 <- nyc$.$date_service_table %>% 
  filter(date == "2018-08-23") %>% select(service_id)
```

(See the `servicepatterns` and `timetable` vignette for more advice on schedule filtering.)

Then we calculate the route frequency in the afternoon rush hour. 

```{r}
nyc <- get_route_frequency(nyc, service_id = services_on_180823, start_hour = 16, end_hour = 19)
```

```{r}
head(nyc$.$routes_frequency)
```

Again, the median headways for the 1 train seem to roughly correspond (1 min off) to those published on [wikipedia entry](https://en.wikipedia.org/wiki/List_of_New_York_City_Subway_services#Train_intervals)

## Mapping Route Frequency

We can also plot the custom headways above along a set of route shapes above.  

We can do this by joining to the simple features routes table.

```{r}
routes_sf_frequencies <- nyc$.$routes_sf %>% 
  inner_join(nyc$.$routes_frequency, by = "route_id") %>% 
          select(route_id,
                 median_headways, 
                 mean_headways, 
                 st_dev_headways, 
                 stop_count)
plot(routes_sf_frequencies)
```
