---
title: "Getting GTFS Data and Mapping with gtfsr"
author: "Danton Noriega <danton.noriega@gmail.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gtfsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, warning=FALSE, message=FALSE, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse = TRUE, cache=FALSE, comment = "#>", fig.width=7.3, fig.height=5)
```

# `gtfsr` (v1.0.3)

`gtfsr` is an R package for easily importing, validating, and mapping transit data that follows the [General Transit Feed Specification (GTFS)](https://developers.google.com/transit/gtfs/) format.

The `gtfsr` package provides functions for converting files following the GTFS format into a single `gtfs` data objects. A `gtfs` object can then be validated for proper data formatting (i.e. if the source data is properly structured and formatted as a GTFS feed) or have any spatial data for stops and routes mapped using `leaflet`. The `gtfsr` package also provides API wrappers for the popular public GTFS feed sharing site [TransitFeeds](https://transitfeeds.com/), allowing users quick, easy access to hundreds of GTFS feeds from within R.


```{r vignette-body, child = 'gtfsr-vignette-body.Rmd'}
```





