---
title: "Introduction"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidytransit-Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidytransit)
library(dplyr)
```
[
![Travis-CI Build
Status](https://travis-ci.com/r-transit/tidytransit.svg?branch=master)](https://travis-ci.com/r-transit/tidytransit)
[![CRAN
status](https://www.r-pkg.org/badges/version/tidytransit?)](https://cran.r-project.org/package=tidytransit)

tidytransit makes it easy to work with transit data by simplifying [General Transit Feed Specification](http://gtfs.org/) data (the standard format for storing transit data) into tidyverse and sf-friendly dataframes. Use it to map existing stops and routes, calculate transit frequencies, and validate transit feeds.

tidytransit is a [fork](https://en.wikipedia.org/wiki/Fork_(software_development)) of [gtfsr](https://github.com/ropensci/gtfsr), published to [CRAN](https://cran.r-project.org/), with frequency calculation functions, and without GTFS-specific interactive cartography features. 

## Installation

This package requires a working installation of [sf](https://github.com/r-spatial/sf#installing). 
```{r, eval=FALSE}
# Once sf is installed, you can install from CRAN with: 
install.packages('tidytransit')

# For the development version from Github:
# install.packages("devtools")
devtools::install_github("r-transit/tidytransit")
```

For some users, `sf` is impractical to install due to system level dependencies. For these users, [`trread`](https://github.com/r-transit/trread) may work better. Its tidytransit without geospatial (GDAL) tools. 

## Usage

### Read GTFS data

```{r}
# Read in GTFS feed
# here we use a feed included in the package, but note that you can read directly from the New York City Metropolitan Transit Authority using the following URL:
# nyc <- read_gtfs("http://web.mta.info/developers/data/nyct/subway/google_transit.zip")

local_gtfs_path <- system.file("extdata", 
                               "google_transit_nyc_subway.zip", 
                               package = "tidytransit")
nyc <- read_gtfs(local_gtfs_path, 
                 local=TRUE,
                 geometry=TRUE,
                 frequency=TRUE)
```

When you add flags for geometry=TRUE and frequency=TRUE, Tidytransit attempts to convert GTFS feeds into simple features dataframes and frequency/headway dataframes upon import of the GTFS data. 

Below we discuss methods and data available for the gtfs object as read by tidytransit. 

### Plot

Perhaps you want to map subway routes and color-code each route by how often trains come. 

```{r}
plot(nyc)
```

### Headways by Route

View the headways along routes as a dataframe. `routes_frequency` is added to the list of gtfs dataframes read in by `read_gtfs` when frequency=TRUE. By default, frequency is calculated for service that happens every weekday from 6 am to 10 pm. See the reference for the `get_route_frequency` function for other options (e.g. weekends, other times of day). 

```{r}
head(nyc$routes_frequency)
```

| route\_id | median\_headways | mean\_headways | st\_dev\_headways | stop\_count |
| :-------- | ---------------: | -------------: | ----------------: | ----------: |
| GS        |                4 |              4 |              0.01 |           4 |
| L         |                4 |              4 |              0.13 |          48 |
| 1         |                5 |              5 |              0.14 |          76 |
| 7         |                5 |              5 |              0.29 |          44 |
| 6         |                6 |              7 |              2.84 |          76 |
| E         |                6 |             23 |             53.01 |          48 |

### Headways by Stop

View the headways at stops. `stops_frequency` is added to the list of gtfs dataframes read in by `read_gtfs`. Again, by default, frequency is calculated for service that happens every weekday from 6 am to 10 pm. See the reference for the `get_stop_frequency` function for other options (e.g. weekends, other times of day). 

```{r}
head(nyc$stops_frequency)
```

    ## # A tibble: 6 x 4
    ## # Groups:   direction_id, stop_id [6]
    ##   direction_id stop_id stop_name             headway
    ##          <int> <chr>   <chr>                   <dbl>
    ## 1            0 902N    Times Sq - 42 St         3.60
    ## 2            1 901S    Grand Central - 42 St    3.60
    ## 3            1 902S    Times Sq - 42 St         3.60
    ## 4            0 901N    Grand Central - 42 St    3.61
    ## 5            0 702N    Mets - Willets Point     3.72
    ## 6            0 707N    Junction Blvd            3.72

### Use the Transitfeeds.com Feedlist 

Included in the tidytransit package is a dataframe with a list of urls, city names, and locations. 

You can browse it as a data frame:

```{r}
head(feedlist_df)
```

or as a map:

```{r}
library(sf)

feedlist_sf <- st_as_sf(feedlist_df,
                        coords=c('loc_lng','loc_lat'),
                        crs=4326)

plot(feedlist_sf, max.plot = 1)
```

### Feed Validation Results

When reading a feed, tidytransit checks it against the GTFS specification, and adds an attribute to the resultant object called `validation_result`, which is a tibble about the files and fields in the GTFS feed and how they compare to the specification. 

You can get this tibble from the metadata about the feed. 

```{r}
validation_result <- attr(nyc, "validation_result")
head(validation_result)
```

