---
title: "trread (gtfs reader)"
author: "Tom Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{trread (gtfs reader)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>",
  message = FALSE,
  error = FALSE,
  warning = FALSE,
  fig.path = "README/README-",
  fig.width=7.3,
  fig.height=5,
  out.width = '100%'
)
```

## Description

`tread` is a package for reading the GTFS data standard into R. It can read directly from URL's or flat files, and does some validation of the data structure against the specification.  

## Goal

This is a highly simplified fork/piece of the [gtfsr](https://github.com/ropensci/gtfsr/) package. 

The goal is to break that package into parts to simplify maintenance.  

## Contributors

Among the many [contributors](https://github.com/ropensci/gtfsr/graphs/contributors), [Danton Noriega](https://github.com/dantonnoriega) wrote much of this package. 

## Usage

Fetch data for a bus system in Accra, Ghana from a URL.  

```{r readme-body, eval=FALSE}
library(trread)
library(dplyr)

accra_gtfs <- import_gtfs("https://github.com/AFDLab4Dev/AccraMobility/raw/master/GTFS/GTFS_Accra.zip")
```

Count and list the number of stops per route. 

```{r, eval=FALSE}
attach(accra_gtfs)

routes_df %>% inner_join(trips_df, by="route_id") %>%
  inner_join(stop_times_df) %>% 
    inner_join(stops_df, by="stop_id") %>% 
      group_by(route_long_name) %>%
        summarise(stop_count=n_distinct(stop_id)) %>%
  arrange(desc(stop_count))
```
