---
title: "Generate a Departure Timetable"
author: "Flavio Poletti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidytransit-Timetable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidytransit)
library(dplyr)
library(ggplot2)
```

In this vignette a departure timetable for a stop is generated and visualised. For some 
analysis it is important to know how and when a single stop is served and workflows to
gather and plot such data can help with this analysis.

## Read GTFS data

We use a feed from the New York Metropolitan Transportation Authority. It is provided as a 
sample feed with tidytransit but you can read it directly from the MTA's website. 

```{r}
local_gtfs_path <- system.file("extdata", "google_transit_nyc_subway.zip", package = "tidytransit")
gtfs <- read_gtfs(local_gtfs_path, local=TRUE)
# gtfs <- read_gtfs("http://web.mta.info/developers/data/nyct/subway/google_transit.zip")
```

## Casting Time Data as hms

In order to work with time fields in the GTFS feed data, we first need to convert them to [`hms::hms()`](https://cran.r-project.org/web/packages/hms/index.html). Times for the whole feed can be converted with `set_hms_times()`. See the `set_hms_times` reference for more detail on how it works. 

```{r}
gtfs <- gtfs %>% set_hms_times()
```

## trip_origin and trip_headsign

To display where a bus (or any public transit vehicle) is headed on a timetable we need the 
column `trip_headsign` in `gtfs$trips`. This is an optional field but our example feed 
provides this information. To display where a vehicle comes from on the timetable we need to 
create a new column in `gtfs$trips` which we'll call `trip_origin`.

```{r}
# get the id of the first stop in the trip's stop sequence
first_stop_id <- gtfs$stop_times %>% 
  group_by(trip_id) %>% 
  summarise(stop_id = stop_id[which.min(stop_sequence)])

# join with the stops table to get the stop_name
first_stop_names <- left_join(first_stop_id, gtfs$stops, by="stop_id")

# rename the first stop_name as trip_origin
trip_origins <- first_stop_names %>% select(trip_id, trip_origin = stop_name)

# join the trip origins back onto the trips
gtfs$trips <- left_join(gtfs$trips, trip_origins, by = "trip_id")
```

```{r}
gtfs$trips %>%
  select(route_id, trip_origin) %>%
  head()
```

In case `trip_headsign` does not exist in the feed it can be generated similarly to 
`trip_origin`:

```{r}
if(!exists("trip_headsign", where = gtfs$trips)) {
  # get the last id of the trip's stop sequence
  trip_headsigns <- gtfs$stop_times %>% 
    group_by(trip_id) %>% 
    summarise(stop_id = stop_id[which.max(stop_sequence)]) %>% 
    left_join(gtfs$stops, by="stop_id") %>% select(trip_id, trip_headsign.computed = stop_name)

  # assign the headsign to the gtfs object 
  gtfs$trips <- left_join(gtfs$trips, trip_headsigns, by = "trip_id")
}
```

## Create A Departure Time Table

To create a departure timetable we first need to find the ids of all stops in the stops table with the same 
same name, as `stop_name` might cover different platforms and thus have multiple stop_ids in the stops table. 

```{r}
stop_ids <- gtfs$stops %>% 
  filter(stop_name == "Times Sq - 42 St") %>% 
  select(stop_id)
```

## Trips departing from stop 

To the selected stop_ids for Time Square, we can join trip columns: `route_id`, `service_id`, `trip_headsign`, and `trip_origin`. Because stop_ids and trips are linked via the stop_times data frame, we do this by joining the stop_ids we've selected to the stop_times data frame and then to the trips data frame. 

```{r}
departures <- stop_ids %>% 
  inner_join(gtfs$stop_times %>% 
               select(trip_id, arrival_time_hms, 
                      departure_time_hms, stop_id), 
             by = "stop_id")

departures <- departures %>% 
  left_join(gtfs$trips %>% 
              select(trip_id, route_id, 
                     service_id, trip_headsign, 
                     trip_origin), 
            by = "trip_id") 
```

## add route info (route_short_name)

Each trip belongs to a route, and the route short name can be added to the departures by joining the trips data frame with `gtfs$routes`.

```{r}
departures <- departures %>% 
  left_join(gtfs$routes %>% 
              select(route_id, 
                     route_short_name), 
            by = "route_id")
```

Now we have a data frame that tells us about the origin, destination, and time at which each train depart from Times Square for every possible schedule of service. 

```{r}
departures %>% 
  select(arrival_time_hms,
         departure_time_hms,
         trip_headsign,trip_origin,
         route_id) %>%
  head() %>%
  knitr::kable()
```

However, we don't know days on which these trips run. Using the service_id column on our calculated departures, and tidytransit's calculated `date_service_table`, we can filter trips to a given date of interest. 

## Preparing a Date Service Table

First, we need to calculate a table that maps calendar dates to service_ids for the entire schedule. To do this, we pipe the feed to the `set_date_service_table` function, which adds a `date_service_table` to the computed tables in the feed. Please the `servicepatterns` vignette for further examples on how to use this table. 

```{r}
gtfs <- gtfs %>% 
  set_hms_times() %>% 
  set_date_service_table()
```

```{r}
head(gtfs$.$date_service_table)
```

## Extract a single day 

Now we are ready to extract the same service table for any given day of the year. 

For example, for August 23rd 2018, a typical weekday, we can filter as follows: 

```{r fig.width=8, fig.height=12}
services_on_180823 <- gtfs$.$date_service_table %>% 
  filter(date == "2018-08-23") %>% select(service_id)

departures_180823 <- departures %>% 
  inner_join(services_on_180823, by = "service_id")
```

How services and trips are set up depends largely on the feed. For an idea how to handle other dates and questions about schedules have a look at the `servicepatterns` vignette. 

```{r}
departures_180823 %>%
  arrange(departure_time_hms, stop_id, route_short_name) %>% 
  select(departure_time_hms, stop_id, route_short_name, trip_headsign) %>% 
  filter(departure_time_hms >= hms::hms(hours = 7)) %>% 
  filter(departure_time_hms < hms::hms(hours = 7, minutes = 10)) %>% 
  knitr::kable()
```

## Simple plot

We'll now plot all departures from Times Square depending on trip_headsign and route. We can
use the route colors provided in the feed.

```{r fig.width=8, fig.height=6}
route_colors <- gtfs$routes %>% select(route_id, route_short_name, route_color)
route_colors$route_color[which(is.na(route_colors$route_color))] <- "454545"
route_colors <- setNames(paste0("#", route_colors$route_color), route_colors$route_short_name)

ggplot(departures_180823) + theme_bw() +
  geom_point(aes(y=trip_headsign, x=departure_time_hms, color = route_short_name), size = 0.2) +
  scale_x_time(breaks = seq(0, max(as.numeric(departures$departure_time_hms)), 3600), labels = scales::time_format("%H:%M")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = route_colors) +
  labs(title = "Departures from Times Square on 08/23/18")
```

Now we plot departures for all stop_ids with the same name, we can separate for different
stop_id. The following plot shows all departures for stop_ids 127N and 127S from 7 to 8 AM.

```{r fig.width=6, fig.height=6}
departures_180823_sub_7to8 <- departures_180823 %>% 
  filter(stop_id %in% c("127N", "127S")) %>% 
  filter(departure_time_hms >= hms::hms(hours = 7) & departure_time_hms <= hms::hms(hour = 8))

ggplot(departures_180823_sub_7to8) + theme_bw() +
  geom_point(aes(y=trip_headsign, x=departure_time_hms, color = route_short_name), size = 1) +
  scale_x_time(breaks = seq(7*3600, 9*3600, 300), labels = scales::time_format("%H:%M")) +
  scale_y_discrete(drop = F) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "bottom") +
  labs(title = "Departures from Times Square on 08/23/18") +
  facet_wrap(~stop_id, ncol = 1)
```

Of course this plot idea can be expanded further. You could also differentiate each route by
direction (using headsign, origin or next/previous stops). Another approach is to calculate 
frequencies and show different levels of service during the day, all depending on the goal 
of your analysis.
